---
title: "Bayesian estimation of allele frequency when genotypes are fully observed"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(cowplot)
```

We are going to assume that we have a site with two alleles: $C$ and $T$.
We do not know the frequency of the $T$ allele, $p_T$.  To estimate it
as a Bayesian, we start with a prior distribution.  You can play with the
values of $\alpha_T$ and $\alpha_C$ here and see how it changes the shape of the
beta prior distribution.

## Assume prior parameters, $\alpha_T$ and $\alpha_C$

```{r, echo=FALSE}
inputPanel(
  numericInput(
    "alpha_1", 
    label = "alpha_T",
    min = 0.00, 
    max = 100, 
    value = 0.5,
    step = 0.1
  ),
  numericInput(
    "alpha_2", 
    label = "alpha_C",
    min = 0.00, 
    max = 100, 
    value = 0.5,
    step = 0.1
  )
)
```

### Here is what the prior distribution looks like:

```{r, echo=FALSE, out.height=100}
renderPlot({
  ggplot() +
    xlim(0, 1) +
    geom_function(
      fun = dbeta,
      args = list(
        shape1 = input$alpha_1,
        shape2 = input$alpha_2
      ),
      n = 1000
    )
  }, 
  height = 150 
)
```


## Set a true allele frequency and simulate some genotypes

We want to estimate the allele freqeuncy from some simulated data.  To
do that, we need to set the true value of $p_T$ and simulate a sample
of genotypes (assuming independence of the two allelic types in an indivdual---essentially
saying we expect Hardy-Weinberg equilibrium) from a population with allele frequency $p_T$. 
Then we pretend we don't actually know $p_T$ and we use the sample to estimate it
by computing the _posterior distribution_ for $p_T$. Recall that the posterior for $p_T$
is a beta distribution: 
$$
p_T \sim \mathrm{Beta}(\alpha_T + \#T, \alpha_C + \#C)
$$

Set $p_T$ and $n$ (the diploid sample size) and hit `Simulate new genotypes` to
simulate a new sample.  The posterior distribution will be computed for each new
sample. (The true value of $p_T$ is at the red vertical line).

Explore what happens when you change the prior.
```{r, echo=FALSE}
inputPanel(
  numericInput(
    "p_T", 
    label = "p_T",
    min = 0.00, 
    max = 1.0, 
    value = 0.3,
    step = 0.01
  ),
  numericInput(
    "n", 
    label = "n",
    min = 1, 
    max = 200, 
    value = 20,
    step = 1
  ),
  actionButton(
    "sim_new", 
    "Simulate new genotypes",
    class = "btn-warning"
  )
)

# simulate new dataset 
genotypes <- eventReactive(
  input$sim_new,
  {
    max_in_row <- 20
    tibble(
      g012 = sample(
        x = 2:0,
        size = input$n,
        prob = c(
          input$p_T^2,
          2 * input$p_T * (1 - input$p_T), 
          (1 - input$p_T) ^ 2
        ),
        replace = TRUE
      )
    ) %>%
      arrange(g012) %>%
      mutate(idx = (1:n()) - 1) %>%
      mutate(
        geno = case_when(
           g012== 2 ~ "TT",
          g012 == 1 ~ "CT",
          g012 == 0 ~ "CC"
        ),
        ypos = -floor(idx / max_in_row),
        xpos = idx %% max_in_row
      )
      
  }
)

renderPlot({
  
  ynum <- max(n_distinct(genotypes()$ypos) - 1, 1)
  ylo <- min(genotypes()$ypos) - 1.0
  yhi <- max(genotypes()$ypos) + 0.2 / ynum
  
  geno_plot <- ggplot(genotypes(), aes(x = xpos)) +
    geom_text(aes(label = geno, colour = geno, y = ypos)) +
    ylim(ylo, yhi) +
    theme_void() +
    theme(
      legend.position = "none"
    ) +
    scale_colour_manual(
      values = c(
        CC = "blue",
        CT = "violet",
        TT = "red"
      )
    ) +
    ggtitle("Sample of Genotypes:")
  
  numT <- sum(genotypes()$g012)
  numC <- nrow(genotypes()) * 2 - numT
  
  post_freq_plot <- ggplot() +
    geom_function(
      fun = dbeta,
      args = list(
        shape1 = numT + input$alpha_1,
        shape2 = numC + input$alpha_2
      ),
      n = 5000
    ) +
    xlim(0,1) +
    xlab("p_T") +
    ylab("Posterior Density") +
    geom_vline(xintercept = input$p_T, colour = "red")
  
  cowplot::plot_grid(
    geno_plot,
    post_freq_plot,
    nrow = 1,
    rel_widths = c(0.65, 0.35)
  )
})
```

